---
title: "Illustration: (Censored) Continuous Response"
format: 
  html:
    html-math-method: mathjax
    toc: true
    number-sections: true
bibliography: transitreg.bib
##vignette: >
##  %\VignetteIndexEntry{transitreg}
##  %\VignetteEngine{quarto::html}
##  %\VignetteDepends{transitreg}
##  %\VignetteKeywords{FIXME}
##  %\VignettePackage{transitreg}
---

```{r setup}
#| echo: false
#| message: false
Sys.setlocale("LC_TIME", "C")
library("transitreg")
data(weatherShannon, package = "transitreg")
```

This example illustrates the application of transition models for continuous
data with additional censoring. To demonstrate the capabilities of
`transitreg()`, the demo data set [`weatherShannon`](/man/weatherShannon.qmd) is used
which provides historical records of daily precipitation sums for a weather
station at Shannon airport, Ireland.

Precipitation is physically limited to $0\,mm\,\text{day}^{-1}$ (days withour
rain) and the positive observations (days with rain) often show a strong
positive skewness which needs to be addressed. While `transitreg()`
automatically addresses for the skewness due to it's flexible assumption about
the response distribution, the excess of zero-observations (dry days) can be
addressed in two ways:

* Define one pseudo-bin which only contains all zero-observations.
* Estimate a left censored transition model; dedicated point-mass
  for all zero-observations.

# Data

The data set [`weatherShannon`](/man/weatherShannon.qmd) contains
$n = `r nrow(weatherShannon)`$ records of daily precipitation sums
for Shannon airport (Ireland) observed between
`r format(min(weatherShannon$date), format = "%B, %Y")` and
`r format(max(weatherShannon$date), format = "%B, %Y")`.
A square-root transformation is often applied to reduce the skewness of
the data.

```{r}
#| plot: true
#| fig.width: 8
#| fig.height: 4
par(mfrow = c(1, 2))
data(weatherShannon, package = "transitreg")
head(weatherShannon)
hist(weatherShannon$rain, freq = FALSE, breaks = 100,
     xlab = "rain", main = "Daily precipitation sums, Shannon airport")
hist(sqrt(weatherShannon$rain), freq = FALSE, breaks = 100,
     xlab = expression(sqrt(rain)), main = "Daily precipitation sums, Shannon airport")
```

For more details, see [`weatherShannon`](man/weatherShannon.qmd).

# Marginal response

For illustration, the first three models are unconditional models
(no covariates), estimating a transition model for the marginal response.

* Model 1a: Untransformed precipitation observations, no censoring.
* Model 1b: Square-root transformed precipitation, no censoring.
* Model 1c: Left censored distribution using square-root tranformed precipitation.


## Model 1a

```{r}
bk1a <- seq(0, 70, by = 0.5)
m1a  <- transitreg(rain ~ theta0 + s(theta),
                   data = weatherShannon,
                   breaks = bk1a)
```

```{r}
x <- seq(0, 70, by = 0.1)
dens1a <- pdf(m1a[1], x)
hist(weatherShannon$rain, freq = FALSE, breaks = bk1a)
lines(x, dens1a, col = 2, lwd = 3)
```

## Model 1b: Square-root transformed response

```{r}
bk1b <- seq(0, sqrt(70), by = 0.5)
m1b  <- transitreg(sqrt(rain) ~ theta0 + s(theta),
                   data = weatherShannon,
                   breaks = bk1b)
```

```{r}
x <- seq(0, sqrt(70), by = 0.01)
dens1b <- pdf(m1b[1], x)
hist(sqrt(weatherShannon$rain), freq = FALSE, breaks = bk1b)
lines(x, dens1b, col = 2, lwd = 3)
```

## Model 1c: Square-root transformed response with censoring

```{r}
bk1c <- seq(0, sqrt(70), by = 0.5)
m1c  <- transitreg(sqrt(rain) ~ theta0 + s(theta),
                   data = weatherShannon,
                   breaks = bk1c,
                   censored = "left")
```

```{r}
x <- seq(0, sqrt(70), by = 0.01)
dens1c <- pdf(m1c[1], x)
hist(sqrt(weatherShannon$rain), freq = FALSE,
     breaks = c(0, 0.01, bk1c[-1L]),
     ylim = c(0, 0.5))
lines(x, dens1c, col = 2, lwd = 3)
```

# Conditional

```{r}
head(weatherShannon)
weatherShannon <- transform(weatherShannon,
                         day = as.integer(format(date, "%j")),
                         year = as.integer(format(date, "%Y")))


```

```{r}
bk2 <- seq(0, sqrt(70), by = 0.5)
m2  <- transitreg(sqrt(rain) ~ theta0 + s(theta) + te(day, theta, k = c(10, 10), bs = c("cc", "tp")),
                  data = weatherShannon,
                  breaks = bk2,
                  censored = "left")
plot(m2)
```


```{r}
#| fig: true
#| out.width: 100%
# Plotting observed data
plot(NA, xlim = c(0, 365),
     ylim = sqrt(c(0, max(weatherShannon$rain))) * 1.05,
     main = "Climatological estimate of daily precipitation sums",
     ylab = expression(sqrt(rain)),
     xlab = "day of the year")
tmp <- lapply(split(weatherShannon, ~ year),
     function(x) lines(x$day, sqrt(x$rain), col = "#00000008"))

# Predict estimated quantiles (climatology)
nd <- data.frame(year = 2024, day = 1:365)
p  <- predict(m2, newdata = nd, type = "quantile",
              prob = c(0.01, 0.25, 0.5, 0.75, 0.99))
matplot(x = nd$day, y = p, type = "l", lty = 1,
        lwd = 2, col = c(4, 2, 1, 2, 4), add = TRUE)
legend("top", bty = "n", legend = colnames(p),
       title = "Estimated quantiles",
       pch = NA, lty = 1, lwd = 2,
       col = c(4, 2, 1, 2, 4), ncol = 5)
```

## Goodness of fit

```{r}
# TODO(R): Currently sqrt(rain) is the issue here
#plot(m2, which = "wormplot")
```
