---
title: "Illustration: Continuous Response"
format: 
  html:
    html-math-method: mathjax
    toc: true
    number-sections: true
bibliography: transitreg.bib
##vignette: >
##  %\VignetteIndexEntry{transitreg}
##  %\VignetteEngine{quarto::html}
##  %\VignetteDepends{transitreg}
##  %\VignetteKeywords{FIXME}
##  %\VignettePackage{transitreg}
---

```{r setup}
#| echo: false
#| message: false
invisible(Sys.setlocale("LC_TIME", "C"))
library("transitreg")
data(weatherShannon, package = "transitreg")
```

This example illustrates the application of transition models for continuous
data using the demo data set [`weatherShannon`](/man/weatherShannon.qmd) 
which consists of historical weather observations at
Shannon airport, Ireland.

# Data

Besides other variables, the data set contains records of daily
maximum temperature (`maxtp`). For computational reasons, only the
last 10 years are used in this illustration.

```{r}
weatherShannon <- subset(weatherShannon, date >= as.Date("2015-01-01"))
```

# Unconditional model

```{r}
m1 <- transitreg(maxtp ~ s(theta), breaks = 201, data = weatherShannon)
```

Estimated distribution:

```{r}
#| fig: true
#| fig.width: 6
#| fig.height: 4
#| out.width: 100%
d1 <- m1[1] # or prodist(m1)[1L]
c(continuous = is_continuous(d1), discrete   = is_discrete(d1))

plot(d1, main = "Estimated density")
plot(d1, cdf = TRUE, tp = TRUE,
     main = "Estimated distribution plus\ntransition probabilities")
```

```{r}
# Weighted mean
mean(d1)
# Quantiles
quantile(d1, probs = seq(0, 1, by = 0.1))
# Probability for days with a maximum temperature < 0 degrees
cdf(d1, x = 0)
# Probability of days with a maximum temperature of ~0 degrees.
# to be precise, the probability for a daily maximum between:
pdf(d1, x = 0)
m1$breaks[which(m1$breaks >= 0)[1L] - 1:0]
```

```{r}
pithist(m1)
#rootogram(m1)
#qqrplot(m1)
```

# Conditional climatological model

```{r}
weatherShannon <- transform(weatherShannon,
                       day  = as.integer(format(date, "%j")),
                       year = as.integer(format(date, "%Y")))

f2 <- maxtp ~ s(theta) + te(day, theta, bs = c("cc", "tp"))
m2 <- transitreg(f2, breaks = 101, data = weatherShannon)
pithist(m2)
rootogram(m2, breaks = 15)

tmp_dates <- as.Date(sprintf("2024-%02d-01", c(1, 4, 7, 10)))
idx <- which(with(weatherShannon, date %in% tmp_dates))

plot(m2[idx], col = 1:4)
legend("left", legend = tmp_dates, col = 1:4, lty = 1, pch = 19)

plot(m2[idx], cdf = TRUE, tp = TRUE, col = 1:4)
legend("left", legend = tmp_dates, col = 1:4, lty = 1, pch = 19)
```

## Prediction

```{r}
nd <- data.frame(day = 111)
tp <- predict(m2, newdata = nd, type = "tp")
tp[, 1:4] # Transition probabilities

d2 <- Transition(tp, breaks = m2$breaks)
print(d2)
plot(d2, cdf = T)
```

```{r}
pithist(m2)
rootogram(m2)
qqrplot(m2)
```

```{r}
bkx <- seq(-20, 50, length.out = 151)
m3 <- transitreg(f2, breaks = bkx, data = weatherShannon)

plot(m3[idx], col = 1:4)
legend("left", legend = tmp_dates, col = 1:4, lty = 1, pch = 19)
```

```{r}
pithist(m3)
rootogram(m3)
qqrplot(m3)
```
